### 2023-06-16 bachelorthesis/25-gyrokinetics.tex
#39: Update headings with bold math
#39: Correction Florian Rath

Because of the large number of particles in the fusion plasma a prediction only using Newton-Maxwell dynamics results in an impossible task for simulation based computations, but this problem can be solved with a statistical approach. For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed. Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often preferred \cite{Garbet2010} which results through evolution of the particle density distribution function in the \textit{Vlasov equation}
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0~.
	\label{eq:vlasov}
\end{gather}
To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be described with the distribution function as follows
\begin{gather}
	n = \int \mathrm{d}\vect{v}\,f(\vect{x}, \vect{v}, t) \qquad j = q \int \mathrm{d}\vect{v}\, v f(\vect{x}, \vect{v}, t)~,
	\label{eq:densitycurrent}
\end{gather}
which are then substituted into the Maxwell equations
\begin{gather}
	\begin{aligned}
		\nabla \cdot \vect{B} &= 0 &\qquad \nabla \times \vect{B} &= \mu_0\left( \sum_\mathrm{species} j + \epsilon_0 \frac{\partial \vect{E}}{\partial t} \right) \\
		\nabla \cdot \vect{E} &= \frac{1}{\epsilon_0} \sum_\mathrm{species} qn &\qquad \nabla \times \vect{E} &= - \frac{\partial \vect{B}}{\partial t}~.
	\end{aligned}
	\label{eq:maxwell}
\end{gather}
The Vlasov equation (\ref{eq:vlasov}) in combination with the Maxwell equations (\ref{eq:densitycurrent}) and (\ref{eq:maxwell}) is the basis of the gyrokinetic model. \cite{Krommes2012}

@@

Dynamics in tokamak plasma is described by spatio-temporal scales which is also referred to as \textit{gyrokinetic ordering}. This seperation compares the tuburbulence frequency $\omega$ and the cyclotron frequency $\omega_\mathrm{c}$ ($\omega/\omega_\mathrm{c} \sim 10^{-3})$ due to strong external magnetic fields which causes a fast gyro motion. The length scales of the turbulence are associated with the wavevector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field. The perpendicular component $k_\perp$ scales with the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the equilibrium density $n_0$, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$. The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak. The parallel component $k_\parallel$ on the other hand scales with the machine size $R$. Experiments show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$. Together this seperation result in the following
\begin{gather}
	\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}
	\label{eq:gyroordering}
\end{gather}
with $\epsilon_\mathrm{g} \ll 1$ which applies to the typical dynamics of a fusion core plasma. \cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows the reduction of parameters in the Vlasov equation (\ref{eq:vlasov}).

@@

For the description of charged particle behaviour in the tokamak device the \textit{gyrocenter coordinates} are used [Fig. \ref{fig:gyrocenterCoords}]. In this set of coordinates the gyrocenter follows the magnetic field with the parallel velocity $v_{\parallel}$. The gyro motion is described together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$.

@@

In the next step the coordinates of the Vlasov equation (\ref{eq:vlasov}) are transformed into the gyrocenter coordinates. For that the particle postion $\vect{x}$ is described through the gyrocenter postion $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$ with the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$. The particle velocity $\vect{v}$ is decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and 
a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$. Through the gyrokinetic ordering a phase space transformation can be performed which is based on the Lie perturbation theory, the Hamiltonian or Lagrangian formalism and an asymtotic expansion in the smallness parameter $\epsilon_\mathrm{g}$. In the lowest order is the magnetic moment $\mu$ as an exact invariant which makes the gyro phase $\zeta$ an ignorable variable. \cite{Garbet2010,Cary1981,Cary1983} Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0~.
	\label{eq:gyrokinetic}
\end{gather}
The time derivative of the magnetic moment $\mu$ is zero because the magnetic moment is an exact invariant.

\subsection{{\boldmath $\delta f$} Approximation and Local Limit}
\label{sub:application}

All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which applies the gyrokinetic equation the $\delta f$ approximation and local limit approximation. For the $\delta f$ approximation the particle density distribution function f is seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}). This is given by
\begin{gather}
	f = f_0 + \delta f \qquad \delta f \sim \rho_* f_0~.
\end{gather}
The distribution function in the equilibrium $f_0$ is set in GKW to the Maxwell distribution function. The perturbed distribution function $\delta f$ is considered to vary perpendicular to the poloidal magnetic field in order of the Larmor radius while the perturbation changes with the system size which results in the ordering
\begin{gather}
	\nabla_\perp(\delta f) = \nabla_\perp (f_0)~.
\end{gather}
The $\delta f$ approximation applies for global and local description where the latter is used in this thesis.
The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities can be considered constant over the radial extent of the simulation volume. Because of the tokamak symmetry all equilibrium constants are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general. \\\bigskip

In addition to this, in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$. Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate. Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}. This results in the coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$. Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formualted as follows
\begin{gather}
	\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S~,
\end{gather}
where $g$ is a function which contains the perturbation $\delta f$ and the equilibrium $f_0$. Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field and the perturbed magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$). The source term $S$ contains the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term. Due to prediodic boundary conditions in the radial direction the radial averaged gradients are fixed in time which is also referred to as \textit{gradient driven} approach. \cite{Peeters2009}

>>>>

Because of the large number of particles in the fusion plasma a prediction on the basis of Newton-Maxwell dynamics results in an impossible task for simulation, but this problem can be solved with a statistical approach. For that the particle density distribution function $f(\vect{x}, \vect{v}, t)$ in the six dimensional phase space $\{\vect{x}$, $\vect{v}\}$ with the particles position $\vect{x}$ and velocity $\vect{v}$ is needed. Because collisions are happening at much smaller frequencies than the characteristic frequenies connected to turbulence, the collisionless model is often preferred \cite{Garbet2010} which results through evolution of the particle density distribution function in the \textit{Vlasov equation}
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{x}} \cdot \frac{\mathrm{d}\vect{x}}{\mathrm{d}t} + \frac{\partial f}{\partial \vect{v}} \cdot \frac{\mathrm{d}\vect{v}}{\mathrm{d}t} = 0~.
	\label{eq:vlasov}
\end{gather}
To obtain a closed system the Maxwell equations with the particle density $n$ and current density $j$ can be described with the distribution function as follows
\begin{gather}
	n = \int \mathrm{d}\vect{v}\,f(\vect{x}, \vect{v}, t) \qquad j = q \int \mathrm{d}\vect{v}\, v f(\vect{x}, \vect{v}, t)~,
	\label{eq:densitycurrent}
\end{gather}
which are then substituted into the Maxwell equations
\begin{gather}
	\begin{aligned}
		\nabla \cdot \vect{B} &= 0 &\qquad \nabla \times \vect{B} &= \mu_0\left( \sum_\mathrm{species} j + \epsilon_0 \frac{\partial \vect{E}}{\partial t} \right) \\
		\nabla \cdot \vect{E} &= \frac{1}{\epsilon_0} \sum_\mathrm{species} qn &\qquad \nabla \times \vect{E} &= - \frac{\partial \vect{B}}{\partial t}~.
	\end{aligned}
	\label{eq:maxwell}
\end{gather}
The Vlasov equation (\ref{eq:vlasov}) in combination with the Maxwell equations (\ref{eq:densitycurrent}) and (\ref{eq:maxwell}) is the basis of the gyrokinetic model. \cite{Krommes2012}

@@

The typical spatio-temporal scales connected to the dynamics in a tokamak plasma allow for the so-called \textit{gyrokinetic ordering} as outlined below. The fast gyromotion is much smaller compared to typical  time scales connected to turbulence ($\omega/\omega_\mathrm{c} \sim 10^{-3})$. The length scales of the turbulence are associated with the wave vector $\vect{k}$ which can be seperated into a perpendicular component $k_\perp = |\vect{k} \times \vect{b}|$ and a parallel component $k_\parallel = |\vect{k} \cdot \vect{b}|$ where $\vect{b}$ is parallel to the poloidal component of the magnetic field. The perpendicular component $k_\perp$ is of the ordre of the thermal Larmor radius $k_\perp^{-1} \sim \rhoth$ which is significantly smaller than the scale on which the equilibrium density $n_0$ varies, which can be expressed with the gradient length $L_\mathrm{n} = |\nabla \ln n_0 |^{-1}$. The gradient length compares with the machine size $R$ which leads to the normalized Larmor radius $\rho_* = \rhoth/R \sim 10^{-3} - 10^{-4}$ with $R$ as major radius of the tokamak. The parallel component $k_\parallel$ on the other hand scales with the machine size $R$. Experiments show that in the core plasma the fluctuation amplitude of the density perturbation $\delta n/n_0$ and magnetic field fluctuation $\delta B/B_0$, with $B_0$ as strength of the magnetic field in equilibrium, is of order $\sim 10^{-4}$. Together this seperation result in the following
\begin{gather}
	\frac{\omega}{\omega_\mathrm{c}} \sim \frac{k_\parallel}{k_\perp} \sim \frac{\rhoth}{L_\mathrm{n}} \sim \frac{\delta n}{n_0} \sim \frac{\delta B}{B_0} \sim \frac{v_\mathrm{d}}{\vth} \sim \epsilon_\mathrm{g}
	\label{eq:gyroordering}
\end{gather}
with $\epsilon_\mathrm{g} \ll 1$ which applies to the typical dynamics of a fusion core plasma. \cite{Brizard2007,Garbet2010} The gyrokinetic ordering allows tallows to formulate reduced governing equations referred to as the gyrokinetic formalism.

@@

For the description of charged particle behaviour in the tokamak device the \textit{guidingcenter coordinates} are used [Fig. \ref{fig:gyrocenterCoords}]. In this set of coordinates the guidingcenter follows the magnetic field with the parallel velocity $v_{\parallel}$. The gyro motion is described together with the magnetic moment $\mu = \frac{1}{2}mv_\perp/\omega_\mathrm{c}$, the gyrocenter $\vect{X}$ and the gyro phase $\zeta$ which gives a parameter set of six quantities $\{\vect{X}, v_{\parallel}, \mu, \zeta\}$. The gyrocenter position $\vect{X}$ is expressed with the particle position $\vect{x}$ and the postion Larmor radius vector $\vect{\rho_\mathrm{L}}$ as $\vect{X} = \vect{x} - \vect{\rho_\mathrm{L}}$.

@@

In the next step the unperturbed Lagrangian is expressed through the guidingcenter coordinates. The particle velocity $\vect{v}$ is decomposed into a parallel component $v_\parallel = \vect{v} \cdot \vect{b}$ and a perpendicular component $v_\perp = |\vect{v} \times \vect{b}|$ which relates to the magnetic moment $\mu$. As final step the Lagrangian is transformed into gyrocenter phase space, based on the so-called Lie perturbation theory. This method allows to eliminate the gyrophase-dependent contributions, rendering the gyrophase $\zeta$ an ignorable coordinate and the magnetic moment $\mu$ an exact constant of motion. \cite{Garbet2010,Cary1981,Cary1983} 
Taking everything into account the new coordinates of the Vlasov equation are $\{\vect{X}, v_{\parallel}, \mu\}$ which transforms the Vlasov equation into the \textit{gyrokinetic equation}, describing the evolution of the gyrocenter distribution function.
\begin{gather}
	\frac{\partial f}{\partial t} + \frac{\partial f}{\partial \vect{X}} \cdot \frac{\mathrm{d}\vect{X}}{\mathrm{d}t} + \frac{\partial f}{\partial v_\parallel} \cdot \frac{\mathrm{d}v_\parallel}{\mathrm{d}t} = 0~.
	\label{eq:gyrokinetic}
\end{gather}
The time derivative of the magnetic moment $\mu$ is zero because the magnetic moment is an exact invariant.

\subsection{$\delta f$ Approximation and Local Limit}
\label{sub:application}

All simulations in this thesis are performed with GKW (Gyro Kinetic Workshop) which solves the gyrokinetic equationin the $\delta f$ approximation and local limit. For the $\delta f$ approximation the particle density distribution function f is seperated into perturbation $\delta f$ and equilibrium $f_0$ (constant in time) as expressed in the gyrokinetic ordering (\ref{eq:gyroordering}). This is given by
\begin{gather}
	f = f_0 + \delta f \qquad \delta f \sim \rho_* f_0~.
\end{gather}
In GKW the equilibrium distribution function $f_0$ is set to the Maxwell distribution function. The perturbed distribution function $\delta f$ is considered to vary perpendicular to the poloidal magnetic field of order of the Larmor radius while the equilibrium changes with the system size which results in the ordering
\begin{gather}
	\nabla_\perp(\delta f) = \nabla_\perp (f_0)~.
\end{gather}
The $\delta f$ approximation applies for global and local description where the latter is used in this thesis.
The \textit{local limit} on the other hand uses the spatial seperation of equilibrium and perturbation and implies that all equilibrium and geometry quantities are considered constant over the radial extent of the simulation volume. Because of the tokamak symmetry all equilibrium quantities are also constant in the toridial direction where the dependency of the equilibrium magnetic field on the poloidal angle is retrained in general. \\\bigskip

In addition to this, in GKW the following coordinates get applied with the parameter $\vect{X} = \{\xcoord, \ycoord, s\}$. Here, $\xcoord$ is the radial coordinate that labels the flux surfaces normalized by the thermal Larmor radius $\rhoth$, $\ycoord$ labels the field lines and is an approximate binormal coordinate. Together with the coordinate $s$ which parameterizes the length along the field lines and is referred to as the parallel coordinate these quantities form the Hamada coordinates \cite{Hamada1958}. This results in the coordinates $\{\xcoord, \ycoord, s, v_\parallel, \mu\}$. Combining the $\delta f$ approximation and the Hamada coordinates the gyrokinetic equation can be formulated as follows
\begin{gather}
	\frac{\partial g}{\partial t} + \vect{v}_\chi \cdot \nabla g + (v_\parallel \vect{b} + \vect{v}_\mathrm{D}) \cdot \nabla(\delta f) - \frac{\mu B}{m}\frac{\vect{B}\cdot\nabla B}{B^2}\frac{\partial (\delta f)}{\partial v_\parallel} = S~,
\end{gather}
where $g$ is a function which contains the perturbation $\delta f$ and the equilibrium $f_0$. Furthermore, $\vect{v}_\mathrm{D}$ and $\vect{v}_\chi$ are drift velocities which are caused by the inhomogeneous magnetic field ($\vect{v}_\mathrm{D}$) and through the $\exb$ drift ($\vect{v}_\chi$). The source term $S$ contains the distribution function in the equilibrium $f_0$ and a correction term for the collisions as well as the energy injection term. Due to prediodic boundary conditions in the radial direction the radial averaged gradients are fixed in time which is also referred to as \textit{gradient driven} approach. \cite{Peeters2009}

### 2023-06-16 bachelorthesis/33-cluster.tex
#39: Correction Florian Rath

The simulation itself will be performed on the \texttt{btrzx1}-cluster of the University Bayreuth. This cluster has a wall time from 24 hours, an total amount of 345 compute nodes with
two AMD Epyc Processors (2nd generation) with 16 cores each and 128 GB of main memory each node and as resource manager \texttt{Slurm}. \cite{btrzx1}

>>>>

The simulation itself will be performed on the \texttt{btrzx1}-cluster of the University Bayreuth. This cluster has a wall time of 24 hours, an total amount of 345 compute nodes with
two AMD Epyc Processors (2nd generation) with 16 cores each and 128 GB of main memory each node and as resource manager \texttt{Slurm}. \cite{btrzx1}

### 2023-06-16 bachelorthesis/34-restartscript.tex
#39: Correction Florian Rath

As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typically run longer than 24 hours, the simulation has to be restarted after the wall time is exceeded. Some simulations need multiple weeks making a regular restart vital to obtaining complete sets of simulation data. The following restart script were written to outsource and simplify the process.

@@

To continue, the status file has different style options, e.g., \texttt{fancy}, \texttt{universal} and \texttt{None} which are displayed in Fig. \ref{fig:statusStyle}. 
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
  ╭───────────────╮    +---------------+    ------------------
  │     fancy     │    |   universal   |           None
  ├───────────────┤    +---------------+    ------------------
  │               │    |               |    
  ╰───────────────╯    +---------------+    ------------------
\end{verbatim}
\captionof{figure}{Styles of output table in status file}
\label{fig:statusStyle}
\end{center}
In addition to that, it dynamically updates the progress of simulation to the end of the status file in form of a progress bar. The progress bar contains the current time step of the simulation in contrast to the required time steps and the progress of the run itself in seconds in relation to the wall time. It also contains the current output from \texttt{squeue}, if the simulation is running or pending, else the important information of the simulation. An example progress bar ist shown in Fig. \ref{fig:statusBar}.
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
 +------------------------------------------------------------+
 |  JOBID  NAME     USER ST    TIME NODES  NODELIST(REASON)   |
 | 970455 3x1.5 bt712347 CG 3:16:01     3  r03n03,r05n[15-16] |
 +------------------------------------------------------------+
 | PROGRESS  [===================>.....]  ( 80%)  20000/25000 |
 | RUN 13    [===>.....................]  ( 13%)  11462/86400 |
 +------------------------------------------------------------+
\end{verbatim}
\captionof{figure}{Progress bar example}
\label{fig:statusBar}
\end{center}
The progress bar displays all important information at one glance. To boost productivity the following command is recommended:
\begin{lstlisting}
cd $DATA 
find . -name $STATUSFILENAME -exec tail -8 {} \;
\end{lstlisting}
which prints all progress bars of every simulation to the data folder. If one want to see the complete status file change \texttt{tail -8} to \texttt{cat}.
An example status file is displayed in the Appendix \ref{append:status}.

>>>>

As stated in Chapter \ref{sec:cluster} the \texttt{btrzx1}-cluster has a wall time of 24 hours but since simulations typically run longer than 24 hours, the simulation has to be restarted after the wall time is exceeded. Some simulations need multiple weeks making a regular restart vital to obtaining complete sets of simulation data. The following restart script was written to outsource and simplify the process.

@@

To continue, the status file has different style options, e.g., \texttt{fancy}, \texttt{universal} and \texttt{None} which are displayed in Fig. \ref{fig:statusStyle}. 
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
  ╭───────────────╮    +---------------+    ------------------
  │     fancy     │    |   universal   |           None
  ├───────────────┤    +---------------+    ------------------
  │               │    |               |    
  ╰───────────────╯    +---------------+    ------------------
\end{verbatim}
\captionof{figure}{Styles of output table in status file}
\label{fig:statusStyle}
\end{center}
In addition to that, it dynamically updates the progress of simulation to the end of the status file in form of a progress bar. The progress bar contains the current time step of the simulation in contrast to the required time steps and the progress of the run itself in seconds in relation to the wall time. It also contains the current output from \texttt{squeue}, if the simulation is running or pending, else the important information of the simulation. An example progress bar ist shown in Fig. \ref{fig:statusBar}.
\begin{center}
\captionsetup{type=figure}
\begin{verbatim}
 +------------------------------------------------------------+
 |  JOBID  NAME     USER ST    TIME NODES  NODELIST(REASON)   |
 | 970455 3x1.5 bt712347 CG 3:16:01     3  r03n03,r05n[15-16] |
 +------------------------------------------------------------+
 | PROGRESS  [===================>.....]  ( 80%)  20000/25000 |
 | RUN 13    [===>.....................]  ( 13%)  11462/86400 |
 +------------------------------------------------------------+
\end{verbatim}
\captionof{figure}{Progress bar example}
\label{fig:statusBar}
\end{center}
The progress bar displays all important information at one glance. To boost productivity the following command is recommended:
\begin{lstlisting}
cd $DATA 
find . -name $STATUSFILENAME -exec tail -8 {} \;
\end{lstlisting}
which prints all progress bars of every simulation to the data folder. If one wants to see the complete status file change \texttt{tail -8} to \texttt{cat}.
An example status file is displayed in the Appendix \ref{append:status}.

### 2023-06-16 bachelorthesis/41-resolution.tex
#39: Correction Florian Rath

At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without fearing numerical dissipation. Numerical dissipatation can therefore result to no formation of zonal flow stuctures, which cause an permanent turbulent state of the simulation. The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation. \\
Because of that, the criteria for the best resolution should be:
\begin{enumerate}
	\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
	\item[\textbf{(2)}] Stability for \textbf{long} time periods 
\end{enumerate}
and the following procedure will be applied for verification:
\begin{enumerate}
	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
	\item Reduce to known minimum number of grid points to verify result in general.
\end{enumerate}

@@

Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab. \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref. \citenum{Peeters2016} in Section IV. Note that, because of selected circular concentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref. \citenum{Peeters2016} for $s$-$\alpha$ geometry. In Fig. \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is simular to the results from Ref. \citenum{Peeters2016} with subdued turbulence after $\sim 3000\,R/\vth$.

@@

As next step to verify the selection of the gradient length $\rlt$ an approach to the finite heat flux threshold was made. As Ref. \citenum{Peeters2016} in Section V concludes the finite heat flux threshold is approximately located at a gradient length of $\rlt = 6.3$ for circular geometry [FIG. 4 of Ref. \citenum{Peeters2016}]. Therefore following parameters were used
\begin{gather*}
	\rlt \in [6.0, 6.3]~.
\end{gather*}
As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occur in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.

@@

In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to
\begin{gather*}
	\Nvpar \in [16, 32, 48, 64]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visible that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period. But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the surpressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set criteria. Due to criteria \textbf{(1)} the selected resolution is
\begin{gather*}
	\boxed{\Nvpar = 48}~.
\end{gather*}
With this number of grid points the time until turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.

@@

To conclude a curial result the number of grid points for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}. In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with both criteria in mind, to the following resolution
\begin{gather*}
	\boxed{\Nmu = 9}~.
\end{gather*}

@@

\subsection{Reduction of Magnetic Field Grid Points $\Ns$}
\label{sub:reduceNs}

In the final step the number of grid points for the magnetic field $\Ns$ gets reduced with the following parameters
\begin{gather*}
	\Ns \in [12, 16]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns12-16-nvpar64-nmu9} can be seen that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the stationary state of the turbulence gets instabil after $\sim 2500\,R/\vth$. This concludes to the following resolution
\begin{gather*}
	\boxed{\Ns = 16}~.
\end{gather*}

@@

Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations is displayed in Tab. \ref{tab:resolution}, with reduced number of grid points for the parallel velocity $\Nvpar$.
\begin{center}
	\captionsetup{type=table}
	\begin{tabular}{l | ccccc | ccccc | c | cc}
		& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
		\hline
		S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
	\end{tabular}
	\captionof{table}{
		Resolution used in this paper: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
	}
	\label{tab:resolution}
\end{center}

>>>>

At the beginning of this work the goal is to estimate the minimal resolution needed to run the simulation without numerical dissipation biasing the physical outcome. Numerical dissipation can therefore result to no formation of zonal flow structures, which cause an permanent turbulent state of the simulation. The goal behind this testing is to reduce \textbf{calculation time} and \textbf{costs} of the simulation. \\
Because of that, the criteria for the best resolution should be:
\begin{enumerate}
	\item[\textbf{(1)}] Subdued turbulence after \textbf{short} time periods
	\item[\textbf{(2)}] Stability for \textbf{long} time periods 
\end{enumerate}
and the following procedure will be applied for verification:
\begin{enumerate}
	\item Reduce only one number of grid points and look if criterias \textbf{(1), (2)} are satisfied
	\item Reduce to known minimum number of grid points to verify result in general.
\end{enumerate}

@@

Starting from the \enquote{Standard resolution with 6th order (S6)} [Tab. \ref{tab:resolutionBenchmark}] the first step is to reproduce the result of Ref. \citenum{Peeters2016} in Section IV. Note that, because of selected circular concentric geometry the used inverse background temperature gradient length is $\rlt = 6.0$ instead of $\rlt = 7.0$ which was used in Section IV of Ref. \citenum{Peeters2016} for $s$-$\alpha$ geometry. In Fig. \ref{fig:eflux-ns16-nvpar64-nmu9} the obtained data is similar to the results from Ref. \citenum{Peeters2016} with subdued turbulence after $\sim 3000\,R/\vth$.

@@

As next step to verify the selection of the gradient length $\rlt$ an approach to the finite heat flux threshold was made. As Ref. \citenum{Peeters2016} in Section V concludes the finite heat flux threshold is approximately located at a gradient length of $\rlt = 6.3$ for circular geometry [FIG. 4 of Ref. \citenum{Peeters2016}]. Therefore, following parameters were used
\begin{gather*}
	\rlt \in [6.0, 6.3]~.
\end{gather*}
As seen in Fig. \ref{fig:eflux-gradient-length} for $\rlt = 6.3$ no suppression of turbulence occurs in the whole time domain, which is in agreement with Ref. \citenum{Peeters2016}.

@@

In the following the number of grid points for the parallel velocity $\Nvpar$ is reduced to
\begin{gather*}
	\Nvpar \in [16, 32, 48, 64]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns16-nvpar16-32-48-64-nmu9} is clearly visible that the resolution with $\Nvpar = 16$ is not suitable for criteria \textbf{(1)} because here the turbulence is not subdued after a long time period. But resolution $\Nvpar = 32$ is not acceptable as well according to criteria \textbf{(2)} since the suppressed turbulence state gain instability after $\sim 3000\,R/\vth$.\\
So to conclude only grid points with $\Nvpar = 48, 64$ are satisfying the set of criteria. Due to criteria \textbf{(1)} the selected resolution is
\begin{gather*}
	\boxed{\Nvpar = 48}~.
\end{gather*}
With this number of grid points the time until turbulence subdued is halfed compared to the benchmark case, i.e., turbulence suppression occur after $\sim 1500\,R/\vth$.

@@

To conclude a final result the number of grid points for the parallel velocity got reduced to $\Nvpar = 48$ according to Chapter \ref{sub:reduceNvpar}. In this case the turbulence does not subdue for the resolution $\Nmu = 6$ which leads, with both criteria in mind, to the following resolution
\begin{gather*}
	\boxed{\Nmu = 9}~.
\end{gather*}

@@

\subsection{Reduction Grid Points along of the Magnetic Field $\Ns$}
\label{sub:reduceNs}

In the final step the number of grid points along the magnetic field $\Ns$ gets reduced with the following parameters
\begin{gather*}
	\Ns \in [12, 16]~.
\end{gather*}
In Fig. \ref{fig:eflux-ns12-16-nvpar64-nmu9} can be seen that the reduction to $\Ns = 12$ does not satisfy the criteria \textbf{(2)} because the stationary state of the turbulence gets instabil after $\sim 2500\,R/\vth$. This concludes to the following resolution
\begin{gather*}
	\boxed{\Ns = 16}~.
\end{gather*}

@@

Together with the results of Chapter \ref{sub:reduceNvpar}, \ref{sub:reduceNmu} and \ref{sub:reduceNs} the final resolution used in the upcoming simulations is displayed in Tab. \ref{tab:resolution}, with reduced number of grid points for the parallel velocity $\Nvpar$.
\begin{center}
	\captionsetup{type=table}
	\begin{tabular}{l | ccccc | ccccc | c | cc}
		& $N_\mathrm{m}$ & $N_x$ & $N_\mathrm{s}$ & $N_{\nu_\parallel}$ & $N_\mu$ & $D$ & $\nu_\mathrm{d}$           & $D_{\nu_\parallel}$ & $D_x$ & $D_y$ & Order & $k_y\rho$ & $k_x\rho$ \\
		\hline
		S6   & 21    & 83    & 16    & 48                  & 9       & 1   & $|\nu_\parallel|$ & 0.2                 & 0.1   & 0.1   & 6     & 1.4       & 2.1       \\
	\end{tabular}
	\captionof{table}{
		Resolution used in this work: Number of toroidal modes $N_m$, number of radial modes $N_x$, number of grid points along the magnetic field $N_s$,number of parallel velocity grid points $N_{\nu_\parallel}$, number of magnetic moment grid points $N_\mu$, dissipation coefficient used in convection along the magnetic field $D$,the velocity in the dissipation scheme $\nu_d$, dissipation coefficient used in the trapping term $D_{\nu_\parallel}$, damping coefficient of radial modes $D_x$, damping coefficient of toroidal modes $D_y$, order of the scheme used for the zonal mode, maximum poloidal wave vector $k_y\rho$, and maximum radial wave vector $k_x\rho$
	}
	\label{tab:resolution}
\end{center}
